Spatio-angular fluorescence microscopy\\ I. Basic theory

### 2018-12-19 papers/basic-theory/paper/paper.tex
Split paper into two.

% Talon's custom libraries
\usepackage{empheq}
\usepackage{bm}

% Talon's custom commands
\setlength\arraycolsep{1.5pt}
\let\originalleft\left
\let\originalright\right
\renewcommand{\left}{\mathopen{}\mathclose\bgroup\originalleft}
\renewcommand{\right}{\aftergroup\egroup\originalright}
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\newcommand{\tensor}[1]{\overset{\text{\tiny$\leftrightarrow$}}{\mb{#1}}}
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\msf}[1]{\mathsf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\mathbf{\mathbf{r}}_o}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rp}{\mathbf{r}_p}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\bs{\nu}}
\providecommand{\taup}{\bs{\tau}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
% \DeclareFontFamily{OT1}{pzc}{}
% \DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
% \DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\newcommand{\eqname}[1]{\tag*{#1}}
\newcommand*\widefbox[1]{\fbox{\hspace{1em}#1\hspace{1em}}}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling an aplanatic microscope imaging an ensemble
  of in-focus fluorescent dipoles as a linear Hilbert-space operator, and we
  show that the operator takes a particularly convenient form when expressed in
  a basis of complex exponentials and spherical harmonics---a form we call the
  dipole spatio-angular transfer function. We demonstrate our formalism by
  analyzing a single-view fluorescence microscope without using the monopole or
  scalar approximations. We show that this imaging system has a spatio-angular
  band limit, and we exploit the band limit to perform efficient simulations.
  Notably, we show that information about the out-of-plane orientation of
  ensembles of in-focus fluorophores is recorded by paraxial fluorescence
  microscopes.
\end{abstract*}

@@

Meanwhile, single-molecule localization microscopy (SMLM) experiments use
spatially sparse fluorescent samples to localize single molecules with precision
that surpasses the diffraction limit. Noise limits the precision of this
localization \cite{foreman2011, chao2016}, and several studies have shown that
model mismatch (e.g. ignoring the effects of vector optics, dipole orientation
\cite{backlund2014}, and dipole rotation \cite{lew2013}) can introduce
localization bias as well. Therefore, the most precise and accurate SMLM
experiments must use an appropriate model and jointly estimate both the position
and orientation of each fluorophore. Several studies have successfully used
vector optics and dipole models to estimate the position and orientation of
single molecules \cite{bohmer2003, lieb2004, toprak2006, aguet2009}, and there
is growing interest in designing optical systems for measuring the position,
orientation, and rotational dynamics of single molecules \cite{agrawal2012,
  backer2014, stallinga2015, zhang2018, zhang2018-2}.

@@

We use the term \textit{monopole} to refer to a model of a fluorophore that
treats it as an isotropic absorber/emitter. Although the term \textit{monopole
  approximation} is not in widespread use, we think it accurately describes the
way many models of fluorescence microscopy treat fluorophores, and we use the
term to distinguish the monopole model from more realistic dipole and
higher-order models. Despite their use in models, electromagnetic monopole
absorber/emitters do not exist in nature. All physical fluorophores absorb and
emit radiation with dipole or higher-order moments, and these moments are always
oriented in space. For a classical mental model of fluorophores we imagine each
dipole as a small oriented antenna (with incoherent absorption and emission
moments) where electrons are constrained to move along a single direction. % With
% this model we can argue classically that monopole radiators do not exist because
% they would violate the conservation of charge.

Although many works implicitly apply the monopole approximation, we have
encountered two explicit justifications: (1) the sample contains many randomly
oriented fluorophores within a resolvable volume or (2) the sample contains
unconstrained rotating fluorophores. While both of these situations yield
monopole-like emitters, neither yields emitters that are perfectly described by
the monopole model. In this work we investigate the dipole model of fluorophores
in detail and find the conditions under which the monopole approximation is
justified.

@@

The large majority of fluorescence microscopes are used to image ensembles of
fluorophores, and most existing modeling techniques make use of the monopole
approximation, the scalar approximation, or both. As discussed above, the
Gibson-Lanni model, the Born-Wolf model, and the Richards-Wolf model are
approximate when applied to fluorescence microscopy data because they only model
monopole emitters. Deconvolution algorithms that use these models may make
biased estimates of fluorophore concentrations since they ignore the dipole
excitation and emission of fluorophores, but we will see that these effects are
small when the sample contains many randomly oriented fluorophores within a
resolvable volume or unconstrained rotating fluorophores.

@@

In the present work we begin to place these three subfields on a common
theoretical footing. First, in section \ref{sec:theory}, we assemble the
mathematical tools for formulating the imaging problem. We start with an
abstract description of fluorescence imaging systems and extend the usual
monopole imaging model to dipoles. Along the way we introduce transfer functions
that will allow us to efficiently analyze spatio-angular microscopes. In section
\ref{sec:results} we analyze and simulate a single-view fluorescence microscope,
and we demonstrate our techniques with four simple phantoms. Finally, in section
\ref{sec:discussion} we discuss the results and their broader implications.

@@

\subsection{Monopole imaging in different bases}\label{sec:monopole}
We start by considering a microscope that images a field of in-focus monopoles
by recording the irradiance on a two-dimensional detector. This section treads
familiar ground, but it serves to establish the concepts and notation that will
be necessary when we extend to the dipole case.

@@

We have been careful to use the term \textit{monopole transfer function} instead
of the commonly-used term \textit{optical transfer function}. We reserve the
term \textit{optical transfer function} for optical systems---the optical
transfer function maps between an input irradiance spectrum and an output
irradiance spectrum in an optical system. We can use optical transfer functions
to model the propagation of light through a microscope, but ultimately we are
always interested in the object, not the light emitted by the object. We will
find the distinction between the optical transfer function and the object
transfer function to be especially valuable when we consider dipoles later in
this section.

\subsubsection{Monopole coherent transfer functions}
Although the Fourier transform can be used to calculate the monopole transfer
function directly from the monopole point spread function, there is a well-known
alternative that exploits coherent transfer functions. The key idea is that the
monopole point spread function can always be written as the absolute square of a
scalar-valued \textit{monopole coherent spread function}, $c(\rd - \ro)$,
defined by
\begin{align}
  |c(\rd - \ro)|^2 = h(\rd - \ro). \label{eq:absquarescalar}
\end{align}
Physically, the monopole coherent spread function corresponds to the
scalar-valued field on the detector with appropriate scaling.

@@

\subsection{Dipole imaging in different bases}\label{sec:dipole}
Now we consider a microscope imaging a field of in-focus dipoles by recording
the irradiance on a two-dimensional detector. A function that assigns a real
number to each point on a plane is not sufficient to specify a field of dipoles
because the dipoles can have different orientations. To represent the object we
need to extend object space to
$\mbb{U} = \mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$---the set of square-integrable
functions on the product space of a plane and a two-dimensional sphere (the
usual sphere embedded in $\mbb{R}^3$). To visualize functions in object space we
imagine a sphere at every point on a plane with a scalar value assigned to every
point on each sphere.

In a delta function basis the object can be represented by a function
$f(\ro, \so)$ called the \textit{dipole density}---the number of dipoles per
unit area and per unit solid angle at position $\ro{}$ and oriented along
$\so{}$. Similar to the monopole case, we model the mapping between the object
and the irradiance in a delta function basis as an integral transform
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd', \ro, \so)f(\ro, \so),
\end{align}
where $h'(\rd', \ro, \so)$ is the irradiance at position $\rd'$ created by a
point source at $\ro$ with orientation $\so$. Notice that we have considered all
possible orientations $\so$ and integrated over the sphere $\mbb{S}^2$. The
dipole density is always symmetric under angular inversion,
$f(\ro, \so) = f(\ro,-\so)$, so we could have chosen to integrate over a
hemisphere and adjusted the definition the dipole density by a factor of two.
For convenience we will continue to integrate over the complete sphere. We note
that all functions in this work with $\so$ as an independent variable are
symmetric under angular inversion, $\so \rightarrow -\so$.

@@

\subsubsection{Dipole spatial transfer function}
We can make our first change of basis by applying the Fourier-convolution
theorem to Eq. \eqref{eq:odpsf}, which yields
\begin{align}
G(\bv) = \int_{\mbb{S}^2}d\so\, H(\bv, \so)F(\bv, \so) \label{eq:odotf},
\end{align}
where we define the \textit{dipole spatial transfer function} as
  \begin{align}
  H(\bv, \so) &= \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}, \so)\, \text{exp}(-2\pi i\mb{r}\cdot\bv),\label{eq:dstf}
  \end{align}
  and the \textit{dipole spatial spectrum} as
  \begin{align}
  F(\bv, \so) &= \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r}, \so)\, \text{exp}(-2\pi i\mb{r}\cdot\bv). 
  \end{align}
  Since the dipole point spread function is normalized and real, we know that
  the dipole spatial transfer function is normalized,
  $\int_{\mbb{S}^2}d\mh{s}\, H(0, \so) = 1$, and conjugate symmetric,
  $H(-\bv, \so) = H^*(\bv, \so)$.

@@

\subsubsection{Dipole angular transfer function}
The spherical harmonics are another set of convenient basis functions that play
the same role as complex exponentials in spatial transfer functions---see
Appendix \ref{sec:sph} for an introduction to the spherical harmonics. We can
change basis from spherical delta functions to spherical harmonics by applying
the generalized Plancherel theorem for spherical functions
\begin{align}
  \ints{2}d\mh{s}\, p(\mh{s})q^*(\mh{s}) = \lmsum P_\ell^m Q_\ell^{m*}, \label{eq:plan}
\end{align}
where $p(\mh{s})$ and $q(\mh{s})$ are arbitrary functions on the sphere,
$P_\ell^m$ and $Q_\ell^m$ are their spherical Fourier transforms defined by
\begin{align}
  P_\ell^m = \int_{\mbb{S}^2}d\mh{s}\, p(\mh{s})Y_\ell^{m*}(\mh{s}),
\end{align}
and $Y_{\ell}^m(\mh{s})$ are the spherical harmonic functions defined in
Appendix \ref{sec:sph}. Equation \eqref{eq:plan} expresses the fact that scalar
products are invariant under a change of basis \cite[Eq.~3.78]{barrett2004}.
The left-hand side of Eq. \eqref{eq:plan} is the scalar product of
$\mbb{L}_2(\mbb{S}^2)$ functions in a delta function basis and the right-hand
side is the scalar product of $\mbb{L}_2(\mbb{S}^2)$ functions in a spherical
harmonic function basis. Applying Eq. \eqref{eq:plan} to Eq. \eqref{eq:odpsf} yields
\begin{align}
  g(\rd) = \lmsum \int_{\mbb{R}^2}d\ro\, H_\ell^m(\rd - \ro)F_\ell^m(\ro), \label{eq:atf-form}
\end{align}
where we have defined the \textit{dipole angular transfer function} as
\begin{align}
  H_\ell^m(\rd - \ro) = \int_{\mbb{S}^2}d\so\, h(\rd - \ro, \so)Y_{\ell}^{m*}(\so),\label{eq:atf-prep} 
\end{align}
and the \textit{dipole angular spectrum} as
\begin{align}
  F_\ell^m(\ro) = \int_{\mbb{S}^2}d\so\, f(\ro, \so)Y_{\ell}^{m*}(\so).
\end{align}
Since the dipole point spread function is normalized and real, we know
that the dipole angular transfer function is normalized,
$\int_{\mbb{R}^2}d\mb{r}\, H_0^0(\mb{r}) = 1$, and conjugate symmetric,
$H_\ell^{-m}(\mb{r}) = (-1)^mH_\ell^{m*}(\mb{r})$.

@@

\subsubsection{Spatio-angular dipole transfer function}
We can arrive at our final basis in two ways: by applying the generalized
Plancherel theorem for spherical functions to Eq. \eqref{eq:odotf} or by applying
the Fourier convolution theorem to Eq. \eqref{eq:atf-form}. We follow the
first path and find that
\begin{align}
G(\bv) = \lmsum \msf{H}_\ell^m(\bv)\msf{F}_\ell^m(\bv) \label{eq:saft},
\end{align}
where we have defined the \textit{dipole spatio-angular transfer function} as
  \begin{align}
  \msf{H}_\ell^m(\bv) &= \int_{\mbb{S}^2}d\so\, H(\bv, \so)Y_\ell^{m*}(\so),
  \end{align}
  and the \textit{dipole spatio-angular spectrum} as
  \begin{align}
  \msf{F}_\ell^m(\bv) &= \int_{\mbb{S}^2}d\so\, F(\bv, \so)Y_\ell^{m*}(\so).
  \end{align}
  Since the dipole point spread function is normalized and real, we know
  that the dipole spatio-angular transfer function is normalized,
  $\msf{H}_0^0(0) = 1$, and conjugate symmetric,
  $\msf{H}_\ell^{-m}(-\bv) = (-1)^m\msf{H}_\ell^{m*}(\bv)$.

  This basis is well suited for simulating and analyzing arbitrary samples
  because it exploits the band limit of the imaging system. We will see the
  advantage explicitly in Section \ref{sec:results} when we calculate a specific
  dipole spatio-angular transfer function and see that it has spatial and
  angular band limits. We note that most single molecule imaging experiments
  are best described in this basis because of the effects of spatial and
  rotational diffusion.

@@

\subsubsection{Dipole coherent transfer functions}
Similar to the monopole case, there is an efficient way to calculate the
transfer functions using coherent transfer functions. The dipole point spread
function can always be written as the absolute square of a vector-valued
function, $\mb{c}(\rd - \ro, \so)$, called the \textit{dipole coherent spread
  function}:
\begin{align}
  |\mb{c}(\rd - \ro, \so)|^2 = h(\rd - \ro, \so). \label{eq:absquare2}
\end{align}
Physically, the dipole coherent spread function corresponds to the vector-valued
electric field on the detector with appropriate scaling. We need a vector-valued
coherent transfer function since the polarization of the field plays a
significant role in dipole imaging, so the dipole point spread function cannot
be written as an absolute square of a scalar-valued function.

@@

We can plug Eq. \eqref{eq:absquare2} into Eq. \eqref{eq:dstf} and use the
autocorrelation theorem to rewrite the dipole spatial transfer function as
\begin{align}
  H(\bv, \so) = \int_{\mbb{R}^2}d\taup\,\mb{C}(\taup, \so)\mb{C}^\dagger(\taup - \bv, \so), 
\end{align}
where we have introduced the \textit{dipole coherent transfer function}
$\mb{C}(\taup, \so)$ as the two-dimensional Fourier transform of the dipole
coherent spread function:
\begin{align}
  \mb{C}(\taup, \so) = \int_{\mbb{R}^2}d\mb{r}\, \mb{c}(\mb{r}, \so)\,\text{exp}[-2\pi i\mb{r}\cdot\taup].
\end{align}
Physically, the dipole coherent transfer function corresponds to the
vector-valued electric field created by a dipole oriented along $\so$ in a
Fourier plane of the detector with appropriate scaling. Similar to the monopole
case, we can calculate the dipole-orientation-dependent fields in a Fourier
plane of the detector, scale appropriately to find the dipole coherent transfer
function, then use the relationships in Fig. \ref{fig:transfer-functions} to
calculate the other transfer functions. Finally, we note that the dipole
coherent transfer function is identical to what Agrawal et. al. call the
\textit{Green's tensor} \cite{agrawal2012} and Novotny and Hecht's
\textit{dyadic point spread function} multiplied by the dipole moment vector
\cite{nov2006}.

>>>>

% Talon's custom libraries
\usepackage{empheq}
\usepackage{bm}
\usepackage[normalem]{ulem}

% Talon's custom commands
\setlength\arraycolsep{1.5pt}
\let\originalleft\left
\let\originalright\right
\renewcommand{\left}{\mathopen{}\mathclose\bgroup\originalleft}
\renewcommand{\right}{\aftergroup\egroup\originalright}
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\newcommand{\tensor}[1]{\overset{\text{\tiny$\leftrightarrow$}}{\mb{#1}}}
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\msf}[1]{\mathsf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\mathbf{\mathbf{r}}_o}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rp}{\mathbf{r}_p}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\bs{\nu}}
\providecommand{\taup}{\bs{\tau}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
\providecommand{\add}[1]{\textcolor{blue}{\uwave{#1}}}
\providecommand{\rem}[1]{\textcolor{red}{\sout{#1}}}
% \DeclareFontFamily{OT1}{pzc}{}
% \DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
% \DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\newcommand{\eqname}[1]{\tag*{#1}}
\newcommand*\widefbox[1]{\fbox{\hspace{1em}#1\hspace{1em}}}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We model an aplanatic microscope imaging an ensemble of in-focus
  fluorescent dipoles as a linear Hilbert-space operator, and we show that the
  operator takes a particularly convenient form when expressed in a basis of
  complex exponentials and spherical harmonics---a form we call the dipole
  spatio-angular transfer function. We discuss the implications of our analysis
  for all quantitative fluorescence microscopy studies and lay out a path
  towards a complete theory. % We demonstrate our formalism by
  % analyzing a single-view fluorescence microscope without using the monopole or
  % scalar approximations. We show that this imaging system has a spatio-angular
  % band limit, and we exploit the band limit to perform efficient simulations.
  % Notably, we show that information about the out-of-plane orientation of
  % ensembles of in-focus fluorophores is recorded by paraxial fluorescence
  % microscopes.
\end{abstract*}

@@

Meanwhile, single-molecule localization microscopy (SMLM) experiments use
spatially sparse fluorescent samples to localize single molecules with precision
that surpasses the diffraction limit. Noise limits the precision of this
localization \cite{foreman2011, chao2016}, and several studies have shown that
model mismatch (e.g. ignoring the effects of vector optics, dipole orientation
\cite{backlund2014}, and dipole rotation \cite{lew2013}) can introduce
localization bias as well. Therefore, the most precise and accurate SMLM
experiments must use an appropriate model and jointly estimate both the position
and orientation of each fluorophore. Several studies have successfully used
vector optics and dipole models to estimate the position and orientation of
single molecules \cite{bohmer2003, lieb2004, toprak2006, aguet2009,
  mortensen2010}, and there is growing interest in designing optical systems for
measuring the position, orientation, and rotational dynamics of single molecules
\cite{agrawal2012, backer2014, stallinga2015, zhang2018, zhang2018-2}.

@@

We use the term \textit{monopole} to refer to a model of a fluorophore that
treats it as an isotropic absorber/emitter. Although the term \textit{monopole
  approximation} is not in widespread use, we think it accurately describes the
way many models of fluorescence microscopy treat fluorophores, and we use the
term to distinguish the monopole model from more realistic dipole and
higher-order models. Despite their use in models, electromagnetic monopole
absorber/emitters do not exist in nature. All physical fluorophores absorb and
emit radiation with dipole or higher-order moments, and these moments are always
oriented in space. For a classical mental model of fluorophores we imagine each
dipole as a small oriented antenna (with incoherent absorption and emission
moments) where electrons are constrained to move along a single direction.

% Although many works implicitly apply the monopole approximation, we have
% encountered two explicit justifications: (1) the sample contains many randomly
% oriented fluorophores within a resolvable volume or (2) the sample contains
% unconstrained rotating fluorophores. While both of these situations yield
% monopole-like emitters, neither yields emitters that are perfectly described by
% the monopole model. In this work we investigate the dipole model of fluorophores
% in detail and find the conditions under which the monopole approximation is
% justified.

@@

The large majority of fluorescence microscopes are used to image ensembles of
fluorophores, and most existing modeling techniques make use of the monopole
approximation, the scalar approximation, or both. As discussed above, the
Gibson-Lanni model, the Born-Wolf model, and the Richards-Wolf model are
approximate when applied to fluorescence microscopy data because they only model
monopole emitters. Deconvolution algorithms that use these models may make
biased estimates of fluorophore concentrations since they ignore the dipole
excitation and emission of fluorophores.

@@

In the present work we begin to place these three subfields on a common
theoretical footing. First, in section \ref{sec:theory}, we consider arbitrary
fluorescence imaging models and lay out a plan for developing a model for
spatio-angular
imaging. % assemble the mathematical tools for formulating the imaging problem.
In section \ref{sec:monopole} we review the familiar monopole imaging model, and
in section \ref{sec:dipole} we extend the model to
dipoles. % Along the way we introduce transfer functions
% that will allow us to efficiently analyze spatio-angular microscopes. In section
% \ref{sec:results} we analyze and simulate a single-view fluorescence microscope,
% and we demonstrate our techniques with four simple phantoms.
Finally, in section \ref{sec:discussion} we discuss the results and their
broader implications.

@@

\section{Monopole imaging}\label{sec:monopole}
We start by considering a microscope that images a field of in-focus monopoles
by recording the irradiance on a two-dimensional detector. This section treads
familiar ground, but it serves to establish the concepts and notation that will
be necessary when we extend to the dipole case.

@@

We have been careful to use the term \textit{monopole transfer function} instead
of the commonly-used term \textit{optical transfer function}. We reserve the
term \textit{optical transfer function} for optical systems---the optical
transfer function maps between an input irradiance spectrum and an output
irradiance spectrum in an optical system. We can use optical transfer functions
to model the propagation of light through a microscope, but ultimately we are
always interested in the object, not the light emitted by the object. We will
find the distinction between the optical transfer function and the object
transfer function to be especially valuable when we consider dipoles in section \ref{sec:dipole}.

\subsection{Monopole coherent transfer functions}
Although the Fourier transform can be used to calculate the monopole transfer
function directly from the monopole point spread function, there is a well-known
alternative that exploits coherent transfer functions. The key idea is that the
monopole point spread function can always be written as the absolute square of a
scalar-valued \textit{monopole coherent spread function}, $c(\rd - \ro)$,
defined by
\begin{align}
  |c(\rd - \ro)|^2 = h(\rd - \ro). \label{eq:absquarescalar}
\end{align}
Physically, the monopole coherent spread function corresponds to the
scalar-valued field on the detector with appropriate scaling.

@@

\section{Dipole imaging}\label{sec:dipole}
Now we consider a microscope imaging a field of in-focus dipoles by recording
the irradiance on a two-dimensional detector. A function that assigns a real
number to each point on a plane is not sufficient to specify a field of dipoles
because the dipoles can have different orientations. To represent the object we
need to extend object space to
$\mbb{U} = \mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$---the set of square-integrable
functions on the product space of a plane and a two-dimensional sphere (the
usual sphere embedded in $\mbb{R}^3$). To visualize functions in object space we
imagine a sphere at every point on a plane with a scalar value assigned to every
point on each sphere.

In a delta function basis the object can be represented by a function
$f(\ro, \so)$ called the \textit{dipole density}---the number of dipoles per
unit area and per unit solid angle at position $\ro{}$ and oriented along
$\so{}$. Similar to the monopole case, we model the mapping between the object
and the irradiance in a delta function basis as an integral transform
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd', \ro, \so)f(\ro, \so),
\end{align}
where $h'(\rd', \ro, \so)$ is the irradiance at position $\rd'$ created by a
point source at $\ro$ with orientation $\so$. Notice that we have considered all
possible orientations $\so$ and integrated over the sphere $\mbb{S}^2$. The
dipole density is always symmetric under angular inversion,
$f(\ro, \so) = f(\ro,-\so)$, so we could have chosen to integrate over a
hemisphere and adjusted the definition of the dipole density by a factor of two.
For convenience we will continue to integrate over the complete sphere. We note
that all functions in this work with $\so$ as an independent variable are
symmetric under angular inversion, $\so \rightarrow -\so$.

@@

\subsection{Dipole spatial transfer function}
We can make our first change of basis by applying the Fourier-convolution
theorem to Eq. \eqref{eq:odpsf}, which yields
\begin{align}
G(\bv) = \int_{\mbb{S}^2}d\so\, H(\bv, \so)F(\bv, \so) \label{eq:odotf},
\end{align}
where we define the \textit{dipole spatial transfer function} as
  \begin{align}
  H(\bv, \so) &= \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}, \so)\, \text{exp}(-2\pi i\mb{r}\cdot\bv),\label{eq:dstf}
  \end{align}
  and the \textit{dipole spatial spectrum} as
  \begin{align}
  F(\bv, \so) &= \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r}, \so)\, \text{exp}(-2\pi i\mb{r}\cdot\bv). 
  \end{align}
  Since the dipole point spread function is normalized and real, we know that
  the dipole spatial transfer function is normalized,
  $\int_{\mbb{S}^2}d\mh{s}\, H(0, \so) = 1$, and conjugate symmetric,
  $H(-\bv, \so) = H^*(\bv, \so)$.

@@

\subsection{Dipole angular transfer function}
The spherical harmonics are another set of convenient basis functions that play
the same role as complex exponentials in spatial transfer functions---see
Appendix \ref{sec:sph} for an introduction to the spherical harmonics. We can
change basis from spherical delta functions to spherical harmonics by applying
the generalized Plancherel theorem for spherical functions
\begin{align}
  \ints{2}d\mh{s}\, p(\mh{s})q^*(\mh{s}) = \lmsum P_\ell^m Q_\ell^{m*}, \label{eq:plan}
\end{align}
where $p(\mh{s})$ and $q(\mh{s})$ are arbitrary functions on the sphere,
$P_\ell^m$ and $Q_\ell^m$ are their spherical Fourier transforms defined by
\begin{align}
  P_\ell^m = \int_{\mbb{S}^2}d\mh{s}\, p(\mh{s})Y_\ell^{m*}(\mh{s}),
\end{align}
and $Y_{\ell}^m(\mh{s})$ are the spherical harmonic functions defined in
Appendix \ref{sec:sph}. Equation \eqref{eq:plan} expresses the fact that scalar
products are invariant under a change of basis \cite[Eq.~3.78]{barrett2004}.
The left-hand side of Eq. \eqref{eq:plan} is the scalar product of
$\mbb{L}_2(\mbb{S}^2)$ functions in a delta function basis and the right-hand
side is the scalar product of $\mbb{L}_2(\mbb{S}^2)$ functions in a spherical
harmonic function basis. Applying Eq. \eqref{eq:plan} to Eq. \eqref{eq:odpsf} yields
\begin{align}
  g(\rd) = \lmsum \int_{\mbb{R}^2}d\ro\, H_\ell^m(\rd - \ro)F_\ell^m(\ro), \label{eq:atf-form}
\end{align}
where we have defined the \textit{dipole angular transfer function} as
\begin{align}
  H_\ell^m(\rd - \ro) = \int_{\mbb{S}^2}d\so\, h(\rd - \ro, \so)Y_{\ell}^{m*}(\so),\label{eq:atf-prep} 
\end{align}
and the \textit{dipole angular spectrum} as
\begin{align}
  F_\ell^m(\ro) = \int_{\mbb{S}^2}d\so\, f(\ro, \so)Y_{\ell}^{m*}(\so).
\end{align}
Since the dipole point spread function is normalized and real, we know
that the dipole angular transfer function is normalized,
$\int_{\mbb{R}^2}d\mb{r}\, H_0^0(\mb{r}) = 1$, and conjugate symmetric,
$H_\ell^{-m}(\mb{r}) = (-1)^mH_\ell^{m*}(\mb{r})$.

@@

\subsection{Spatio-angular dipole transfer function}
We can arrive at our final basis in two ways: by applying the generalized
Plancherel theorem for spherical functions to Eq. \eqref{eq:odotf} or by applying
the Fourier convolution theorem to Eq. \eqref{eq:atf-form}. We follow the
first path and find that
\begin{align}
G(\bv) = \lmsum \msf{H}_\ell^m(\bv)\msf{F}_\ell^m(\bv) \label{eq:saft},
\end{align}
where we have defined the \textit{dipole spatio-angular transfer function} as
  \begin{align}
  \msf{H}_\ell^m(\bv) &= \int_{\mbb{S}^2}d\so\, H(\bv, \so)Y_\ell^{m*}(\so),
  \end{align}
  and the \textit{dipole spatio-angular spectrum} as
  \begin{align}
  \msf{F}_\ell^m(\bv) &= \int_{\mbb{S}^2}d\so\, F(\bv, \so)Y_\ell^{m*}(\so).
  \end{align}
  Since the dipole point spread function is normalized and real, we know
  that the dipole spatio-angular transfer function is normalized,
  $\msf{H}_0^0(0) = 1$, and conjugate symmetric,
  $\msf{H}_\ell^{-m}(-\bv) = (-1)^m\msf{H}_\ell^{m*}(\bv)$.

  This basis is well suited for simulating and analyzing arbitrary samples
  because it exploits the band limit of the imaging system. % We will see the
  % advantage explicitly in Section \ref{sec:results} when we calculate a specific
  % dipole spatio-angular transfer function and see that it has spatial and
  % angular band limits. 
  We note that most single molecule imaging experiments
  are best described in this basis because of the effects of spatial and
  rotational diffusion.

@@

\subsection{Dipole coherent transfer functions}
Similar to the monopole case, there is an efficient way to calculate the
transfer functions using coherent transfer functions. The dipole point spread
function can always be written as the absolute square of a vector-valued
function, $\mb{c}(\rd - \ro, \so)$, called the \textit{dipole coherent spread
  function}:
\begin{align}
  |\mb{c}(\rd - \ro, \so)|^2 = h(\rd - \ro, \so). \label{eq:absquare2}
\end{align}
Physically, the dipole coherent spread function corresponds to the vector-valued
electric field on the detector with appropriate scaling. We need a vector-valued
coherent transfer function since the polarization of the field plays a
significant role in dipole imaging, so the dipole point spread function cannot
be written as an absolute square of a scalar-valued function.

@@

We can plug Eq. \eqref{eq:absquare2} into Eq. \eqref{eq:dstf} and use the
autocorrelation theorem to rewrite the dipole spatial transfer function as
\begin{align}
  H(\bv, \so) = \int_{\mbb{R}^2}d\taup\,\mb{C}(\taup, \so)\mb{C}^\dagger(\taup - \bv, \so), 
\end{align}
where we have introduced the \textit{dipole coherent transfer function}
$\mb{C}(\taup, \so)$ as the two-dimensional Fourier transform of the dipole
coherent spread function:
\begin{align}
  \mb{C}(\taup, \so) = \int_{\mbb{R}^2}d\mb{r}\, \mb{c}(\mb{r}, \so)\,\text{exp}[-2\pi i\mb{r}\cdot\taup].
\end{align}
Physically, the dipole coherent transfer function corresponds to the
vector-valued electric field created by a dipole oriented along $\so$ in a
Fourier plane of the detector with appropriate scaling. Similar to the monopole
case, we can calculate the dipole-orientation-dependent fields in a Fourier
plane of the detector, scale appropriately to find the dipole coherent transfer
function, then use the relationships in Fig. \ref{fig:transfer-functions} to
calculate the other transfer functions. Finally, we note that the dipole
coherent transfer function is identical (up to scaling factors) to what
Agrawal et. al. call the \textit{Green's tensor} \cite{agrawal2012} and Novotny
and Hecht's \textit{dyadic point spread function} multiplied by the dipole
moment vector \cite{nov2006}.

% \section{Results}\label{sec:results}
% In this section we begin to apply the tools we developed in the previous section
% to analyze a specific fluorescence microscope. During our initial discussion we
% only required the optical system to be aplanatic so that we could model it as a
% shift-invariant magnifier. Now we will be more specific and consider an
% aplanatic optical system in a \textit{$\mathit{4}f$ configuration} with an
% arbitrary first lens (the objective lens) and a \textit{paraxial second lens}
% (the tube lens) as shown in Fig. \ref{fig:schematic}. A lens can be considered
% paraxial if the angle $\alpha$ between the optical axis of the lens and the
% marginal ray is small enough that $\sin\alpha \approx \alpha$. As a rule of
% thumb, non-paraxial effects only become significant when the numerical aperture
% of a lens exceeds 0.7 \cite[ch.~6]{gu2000}, but this is only a rough
% guideline. Commercial microscopes with infinity-corrected objectives can almost
% always can be modeled by considering the tube lens as paraxial.

% We start by defining \textit{pupil functions} for both monopole and dipole
% imaging systems, and we explicitly relate the pupil functions to the previously
% defined coherent transfer functions. After establishing the link between
% physical calculations and the transfer functions, we calculate the monopole and
% dipole transfer functions for imaging systems with a paraxial objective, and we
% demonstrate our approach by simulating simple phantoms.
